\papertitle

### 2024-09-13 src/tex/ms.tex
Ben edits

% Define document class
\documentclass[reprint,superscriptaddress,nobibnotes,amsmath,amssymb,aps,prx,hidelinks,linenumbers]{revtex4-2}

% Import relevant packages
\usepackage{showyourwork}
\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{bm}
\usepackage{multirow}
\usepackage[separate-uncertainty=true]{siunitx}
\usepackage{amsmath}
\usepackage[shortlabels]{enumitem}
\usepackage[page]{appendix}
\usepackage{xfrac}
\usepackage{printlen}
\usepackage[version=4]{mhchem}
\usepackage{blindtext}
% \usepackage[hidelinks]{hyperref}
\sisetup{
    list-units      = brackets,
    range-units     = brackets,
    range-phrase    = -,
    list-pair-separator= {, },
    list-separator  = {, },
    list-final-separator = {, }
    }
\renewcommand\appendixname{}
\newcommand{\papertitle}{Accurate Estimation of Diffusion Coefficients and their Uncertainties from Computer Simulation}
% \newcommand{\oMSD}{\ensuremath{\overline{\bm{s}^2}}}
% \newcommand{\oMSDs}[1]{\ensuremath{\overline{s^2}}(#1)}
% \newcommand{\oMSDi}{\ensuremath{\overline{s^2}_i}}
% \newcommand{\oMSDj}{\ensuremath{\overline{s^2}_j}}

@@

\graphicspath{{figures/}}

% Redefine \maketitle so that it can be used twice (for supplementary)
\makeatletter
\def\maketitle{
\@author@finish
\title@column\titleblock@produce
\suppressfloats[t]}
\makeatother

% Begin!
\begin{document}

\let\oldaddcontentsline\addcontentsline% Store \addcontentsline
\renewcommand{\addcontentsline}[3]{}

% Title
\title{\papertitle}

% Author list
\author{Andrew R. McCluskey}
  \email{andrew.mccluskey@bristol.ac.uk}
  \affiliation{School of Chemistry, University of Bristol, Cantock's Close, Bristol, BS8 1TS, United Kingdom}
  \affiliation{European Spallation Source ERIC, Ole Maaløes vej 3, 2200 København N, DK}
\author{Samuel W. Coles}
  \affiliation{Department of Chemistry, University of Bath, Claverton Down, Bath, BA2 7AY, UK}
  \affiliation{The Faraday Institution, Quad One, Harwell Science and Innovation Campus, Didcot, OX11 0RA, UK}
\author{Benjamin J. Morgan}
  \email{b.j.morgan@bath.ac.uk}
  \affiliation{Department of Chemistry, University of Bath, Claverton Down, Bath, BA2 7AY, UK}
  \affiliation{The Faraday Institution, Quad One, Harwell Science and Innovation Campus, Didcot, OX11 0RA, UK}

% Abstract
\begin{abstract}
    Self-diffusion coefficients, $\D$, are routinely estimated from molecular dynamics simulations by fitting a linear model to the observed mean-squared displacements (MSDs) of mobile species.
    MSDs derived from simulation suffer from statistical noise, which introduces uncertainty in the resulting estimate of $\D$.
    An optimal scheme for estimating $\D$ will minimise this uncertainty, i.e., will have high statistical efficiency, and will give an accurate estimate of the uncertainty itself.
    We present a scheme for estimating $\D$ from a single simulation trajectory with high statistical efficiency and accurately estimating the uncertainty in the predicted value.
    The statistical distribution of MSDs observable from a given simulation is modelled as a multivariate normal distribution using an analytical covariance matrix for an equivalent system of freely diffusing particles, which we parameterise from the available simulation data. 
    We then perform Bayesian regression to sample the distribution of linear models that are compatible with this model multivariate normal distribution, to obtain a statistically efficient estimate of $D^*$ and an accurate estimate of the associated statistical uncertainty.
\end{abstract}

@@

@@

Mass transport is a fundamental physical process that is central to our understanding of fluids~\cite{sendner_interfacial_2009,shimizu_structural_2015,ghoufi_ultrafast_2016} and plays a critical role in biochemical systems~\cite{maccmmon_dynamics_1977,robertson_diffusion_2006}, and in solid-state devices such as batteries, fuel cells, and chemical sensors~\cite{eames_ionic_2015,morgan_understanding_2011,walsh_taking_2018}.
Molecular dynamics simulations are widely used to study microscopic transport processes, as they give direct insight into atomic-scale transport mechanisms and can be used to calculate macroscopic transport coefficients~\cite{morgan_relationships_2014,morgan_mechanistic_2021,poletayev_defect_2022,klepis_long_2009,wang_application_2011,zelovich_hydroxide_2019}.
These transport coefficients are formally defined in terms of ensemble averages.
Dynamical simulations, however, sample the full ensemble space stochastically, and parameters derived from simulation data are therefore only estimates of the true parameter of interest.
The statistical uncertainty associated with such estimates depends on the details of the simulation---e.g., size and timescale---and on the choice of estimation method.
An optimal estimation method will minimise the uncertainty in the computed quantity, i.e., it will have high statistical efficiency, and will also allow this uncertainty to be accurately estimated.

One commonly used parameter for quantifying atomic-scale mass transport is the self-diffusion coefficient, $\D$, which describes diffusion in the absence of a chemical potential gradient.
$\D$ is related to the ensemble-average mean squared displacement (MSD), $\MSD{t}$, via the Einstein relation~\cite{einstein_uber_1905,helfand_transport_1960},
%
\begin{equation}
    \D = \lim_{t\to\infty}\frac{\MSD{t}}{6t},
    \label{equ:einstein}
\end{equation}
%
where $t$ is elapsed time.
Because numerical simulations are finite in both time and space, MSDs obtained from simulation data always deviate from the true ensemble average MSD.
One can, however, compute an estimate of the self-diffusion coefficient, $\Dest$, by fitting a linear model to the observed MSD and using the gradient of this fitted model in place of $\MSD{t} / t$ in Eqn.~\ref{equ:einstein}.

The simplest approach to fitting a linear model to MSD data from simulation is ordinary least squares regression (OLS).
OLS gives analytical expressions for the ``best fit'' regression coefficients (the slope and intercept) and their respective uncertainties, making it easy to implement and quick to perform.
This procedure, however, is appropriate only for data that are both statistically independent and identically distributed.
Neither of these conditions hold for MSD data obtained from simulation, which instead are serially correlated and usually have unequal variances.
As a consequence, OLS is statistically inefficient, giving a relatively large statistical uncertainty in $\Dest$.
Furthermore, using the textbook OLS expression for the uncertainty in $\Dest$ significantly underestimates the true uncertainty in this estimate.
This underestimated uncertainty may cause overconfidence in the accuracy of values of $\D$ estimated using OLS, and using these data in downstream analyses may result in faulty inferences.
While the uncertainty associated with OLS estimates of $\D$ can, in principle, be accurately estimated by directly sampling over multiple repeated simulations, this approach greatly increases the total computational cost and therefore is often not practical.

\begin{figure*}
    \centering
    % \resizebox{\columnwidth}{!}{\includegraphics*{figure1.pdf}}
    \includegraphics[width=\textwidth]{schematic.pdf}
    \caption{
        A schematic diagram of the Bayesian regression method described in this work, running from our simulations through the sampling of displacement, variance and covariance estimation and final the sampling process to give the marginal posterior distribution for $\D$.}
    \label{fig:schematic}
\end{figure*}

Here, we describe an approximate Bayesian regression method for estimating $\D$ with near-maximal statistical efficiency while also accurately estimating the corresponding statistical uncertainty, using data from a single simulation (Fig.~\ref{fig:schematic}).
We model the statistical population of simulation MSDs as a multivariate normal distribution, using an analytical covariance matrix derived for an equivalent system of freely diffusing particles, with this covariance matrix parameterised from the observed simulation data.
We then use Markov-chain Monte Carlo to sample the posterior distribution of linear models compatible with this multivariate normal model.
The resulting posterior distribution provides an efficient estimate for $\D$ and allows the associated statistical uncertainty in $\Dest$ to be accurately quantified.
This method is implemented in the open-source Python package \textsc{kinisi}~\cite{mccluskey_kinisi_2022}.

\section{Results}
\subsection{Background}

For a simulation of equivalent particles, the observed mean squared displacement as a function of time, $\oMSDs{t}$, can be computed as an average over equivalent particles and time origins:
%
\begin{equation}
  \oMSDs{t} = \frac{1}{N(t)}\sum^{N(t)}_{j=1}{\left[\Delta\mathbf{r}_j(t)\right]}^2,
  \label{equ:observed_msd}
\end{equation}
%
where $N(t)$ is the total number of observed squared-displacements at time $t$.
The resulting observed MSD is a vector, $\oMSD$, with individual elements $\oMSDi$.
Each element of this vector differs from the true ensemble-average MSD for that time by some unknown amount.
Fitting a linear model to $\oMSD$ gives an estimated self-diffusion coefficient, $\Dest$, which again differs from the true self-diffusion coefficient, $\D$, by some unknown amount.

Performing repeated simulations starting from different random seeds or with different histories will produce a set of replica trajectories, where each trajectory gives a different, statistically equivalent, observed MSD.
The set of all possible replica trajectories defines a population of hypothetical observed MSDs, and the MSD obtained from any one trajectory can be considered a random sample, $\bm{X}$, drawn from the multivariate probability distribution that describes this population, i.e, $\bm{X} \sim \prob{\oMSD}$.
Each potential MSD sample could, in principle, be fitted to a linear model to obtain a corresponding estimate for the self-diffusion coefficient; $\bm{X} \mapsto \Dest$.
The population of all such estimates therefore defines a probability distribution $\prob{\Dest}$.
The estimated diffusion coefficient obtained from a single simulation corresponds to a random sample drawn from this distribution, while the uncertainty in $\Dest$ is described by the shape of the full distribution $\prob{\Dest}$.

@@

For data that are both statistically independent and identically normally distributed, ordinary least squares regression (OLS) is unbiased and statistically efficient, and gives accurate estimates of the uncertainties in the resulting regression coefficients.
MSD data obtained from simulation, however, are neither statistically independent nor identically distributed.
The variances, $\var{\oMSDi}$, are correlated, since the displacement of each particle at time $t+\Delta t$ is necessarily similar to its displacement at time $t$, and hence, $\oMSDs{t}$ is similar to $\oMSDs{t+\Delta t}$.
These variances are also typically unequal---the data are heteroscedastic \cite{smith_random_1996,he_statistical_2018,UslerEtAl_JComputChem2023}. 
Because the key assumptions of the OLS method are not valid for MSD data, OLS gives statistically inefficient estimates of $\D$, while the estimated regression uncertainties obtained from the standard OLS statistical formulae significantly underestimate the true uncertainty in $\prob{\Dest_\mathrm{OLS}}$ (Fig.~\ref{fig:glswlsols}a).

\begin{figure}
    \centering
    % \resizebox{\columnwidth}{!}{\includegraphics*{figure1.pdf}}
    \includegraphics[width=\columnwidth]{glswlsols.pdf}
    \script{glswlsols.py}
    \caption{
        Example distributions of estimated self-diffusion coefficients, $\Dest$, calculated using (a) ordinary least squares (OLS), (b) weighted least squares (WLS), and (c) generalised least squares (GLS),
        from MSD data from \num{4096} individual simulations of \num{128} particles undergoing a \SI{128}{step} 3D lattice random walk, with a step size chosen so that the true diffusion coefficient $\D = 1$.
        In each panel, the grey curve shows the best-fit normal distribution for the simulation data, the upper horizontal bar shows the standard deviation of this distribution, and the lower horizontal bar shows the average estimated standard distribution given by the analytical expression for $\sigma[\prob{\Dest}]$ for each regression method.}
    \label{fig:glswlsols}
\end{figure}

Some improvement can be made by using weighted least squares (WLS) (Fig.~\ref{fig:glswlsols}b), where the residual for each observed MSD value is weighted by the reciprocal of its variance, $1/(\var{\oMSDi})$.
Like OLS, WLS is an unbiased estimator, and for heteroscedastic data it has higher statistical efficiency than OLS.
WLS still disregards correlations in $\oMSD$, however, and is therefore statistically inefficient, while the WLS estimated uncertainties for the regression coefficients still underestimate the true uncertainty in $\prob{\Dest_\mathrm{WLS}}$.

To optimally estimate the true ensemble-average MSD, and hence $\D$, from simulation data, it is necessary to account for both the changing variance and correlation structure of $\oMSD$.
Within the framework of linear regression, this can be achieved using generalised least squares (GLS).
GLS gives estimated regression coefficients, $\widehat{\beta}$, via
%
\begin{equation}
    \widehat{\beta} = \left(\modelmatrix^{\top}\mathbf{\Sigma}^{-1}\modelmatrix\right)^{-1}\modelmatrix^{\top}\mathbf{\Sigma}^{-1}\oMSD,
    \label{equ:gls}
\end{equation} 
%
where $\modelmatrix$ is the model matrix $\begin{bmatrix}\mathbf{1} & \bm{t}\end{bmatrix}$, with $\bm{t}$ the vector of observed times, and $\mathbf{\Sigma}$ is the covariance matrix for the observed MSD values.
For correlated heteroscedastic data, such as MSD data, GLS offers the theoretical maximum statistical efficiency---it achieves the Cram\'er--Rao bound~\cite{cramer_mathematical_1946,rao_information_1945,rao_selected_1994,darmois_sur_1945,aitken_on_1942}---and provides accurate analytical estimates of the uncertainty in the predicted regression coefficients (Fig.~\ref{fig:glswlsols}c).

An alternative method for estimating the ensemble-average MSD, and thus $\Dest$, from simulation data is Bayesian regression.
Like GLS, Bayesian regression can take into account both the changing variance and the correlation structure inherent in the data.
Rather than providing a singular ``best-fit'' estimate like GLS, Bayesian regression produces a posterior probability distribution for the regression coefficients.
The mean of this distribution serves as the point estimate of the coefficients and, in the absence of additional prior information, is equivalent to the GLS estimate, while the spread of the distribution quantifies the uncertainty in these estimates.
For data that is both heteroscedastic and correlated, such as MSD data from simulations, Bayesian regression, like GLS, is formally fully statistically efficient.

To estimate $\D$ from some observed MSD data, $\oMSD$, using Bayesian regression, we compute the posterior probability distribution $\prob{\model|\oMSD}$ for a linear model $\model = 6\D \bm{t} + c$, where $\D$ and $c$ are parameters to be estimated.
This posterior distribution is described by Bayes' theorem,
%
\begin{equation}
    \prob{\model|\oMSD} = \frac{\prob{{\oMSD|\model}}\prob{\model}}{\prob{\oMSD}},
    \label{equ:bayes}
\end{equation}
%
where $\prob{\oMSD|\model}$ is the probability of observing data $\oMSD$ given model $\model$, often described as the ``likelihood'', and $\prob{\oMSD}$ is the marginal probability of the observed data $\oMSD$.
Integrating over $\prob{\model|\oMSD}$ with respect to $c$ yields the marginal posterior distribution $\prob{\D|\oMSD}$, from which the best point-estimate $\Dest$ and distribution variance $\varest{\Dest}$ can be computed.

Given a sufficiently large number of observed squared displacements at each time $t$, the central limit theorem applies, and $\oMSD$ can be considered a sample from a multivariate normal distribution with log-likelihood
%
\begin{equation}
    \begin{aligned}{}
      \ln \prob{\oMSD|\model} = -\frac{1}{2}\big[ & \ln(\left|\mathbf{\Sigma}\right|) + {(\oMSD - \model)}^{\!\top}\mathbf{\Sigma}^{-1}(\oMSD - \model) \\ 
      & + k \ln(2\pi)\big],
    \end{aligned}
    \label{equ:loglike}
\end{equation}
%
where $\mathbf{\Sigma}$ is the observed MSD covariance matrix and $k$ is the length of the vector $\oMSD$, i.e., the number of time intervals for which we have observed MSD data.
Providing this likelihood function can be calculated, we can compute the posterior distribution $\prob{\model | \oMSD}$, which gives an optimally efficient point-estimate for $\D$ and a complete description of the associated uncertainty in $\Dest$.

\subsection{Approximating $\bm{\Sigma}$ from simulation data}

For Bayesian regression and GLS, we require the covariance matrix for the observed MSD, $\mathbf{\Sigma}$, which is generally unknown.
To proceed, we replace $\mathbf{\Sigma}$ with a model covariance matrix, $\mathbf{\Sigma^\prime}$, with a known analytical form, that we parameterise from the available simulation data.
Providing the correlation structure of $\mathbf{\Sigma^\prime}$ is similar to that of $\mathbf{\Sigma}$, this model correlation matrix can be used in approximate Bayesian or GLS schemes to estimate the ensemble-average MSD, and hence $\D$, with high efficiency and accurate estimated uncertainties.

>>>>

\documentclass[reprint,superscriptaddress,nobibnotes,amsmath,amssymb,aps,prx,hidelinks]{revtex4-2}

\usepackage{showyourwork}
\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{bm}
\usepackage{multirow}
\usepackage[separate-uncertainty=true]{siunitx}
\usepackage{amsmath}
\usepackage[shortlabels]{enumitem}
\usepackage[page]{appendix}
\usepackage{xfrac}
\usepackage{printlen}
\usepackage[version=4]{mhchem}
\usepackage{blindtext}
\usepackage[noabbrev,nameinlink,capitalize]{cleveref}
\crefname{equation}{Eqn.}{Eqns.}
\crefformat{equation}{Eqn.~#2#1#3}
\crefrangeformat{equation}{Eqns.~#3#1#4 to~#5#2#6}
\crefmultiformat{equation}{Eqns.~#2#1#3}{ and~#2#1#3}{, #2#1#3}{ and~#2#1#3}
\crefname{figure}{Fig.}{Figs.}

\sisetup{
    list-units      = brackets,
    range-units     = brackets,
    range-phrase    = -,
    list-pair-separator= {, },
    list-separator  = {, },
    list-final-separator = {, }
    }
\renewcommand\appendixname{}
\newcommand{\papertitle}{Accurate Estimation of Diffusion Coefficients and their Uncertainties from Computer Simulation}

@@

\newcommand{\todo}[1]{\color{red}[#1]\color{black}}

\hyphenation{metalwalls}

\makeatletter
\def\maketitle{
\@author@finish
\title@column\titleblock@produce
\suppressfloats[t]}
\makeatother

\begin{document}

\let\oldaddcontentsline\addcontentsline
\renewcommand{\addcontentsline}[3]{}

\title{\papertitle}

\author{Andrew R. McCluskey}
\email{andrew.mccluskey@bristol.ac.uk}
  \affiliation{Centre for Computational Chemistry, School of Chemistry, University of Bristol, Cantock's Close, Bristol, BS8 1TS, UK.}
  \affiliation{European Spallation Source ERIC, Data Management and Software Centre, Asmussens Allé 305, DK-2800 Kongens Lyngby, DK.}
  \affiliation{Diamond Light Source, Harwell Campus, Didcot, OX11 0DE, UK.}
\author{Samuel W. Coles}
  \affiliation{Department of Chemistry, University of Bath, Claverton Down, Bath, BA2 7AY, UK}
  \affiliation{The Faraday Institution, Quad One, Harwell Science and Innovation Campus, Didcot, OX11 0RA, UK}
\author{Benjamin J. Morgan}
  \email{b.j.morgan@bath.ac.uk}
  \affiliation{Department of Chemistry, University of Bath, Claverton Down, Bath, BA2 7AY, UK}
  \affiliation{The Faraday Institution, Quad One, Harwell Science and Innovation Campus, Didcot, OX11 0RA, UK}

\begin{abstract}
    Self-diffusion coefficients, $\D$, are routinely estimated from molecular dynamics simulations by fitting a linear model to the observed mean-squared displacements (MSDs) of mobile species.
    MSDs derived from simulation exhibit statistical noise that causes uncertainty in the resulting estimate of $\D$.
    An optimal scheme for estimating $\D$ minimises this uncertainty, i.e., it will have high statistical efficiency, and also gives an accurate estimate of the uncertainty itself.
    We present a scheme for estimating $\D$ from a single simulation trajectory with high statistical efficiency and accurately estimating the uncertainty in the predicted value.
    The statistical distribution of MSDs observable from a given simulation is modelled as a multivariate normal distribution using an analytical covariance matrix for an equivalent system of freely diffusing particles, which we parameterise from the available simulation data. 
    We use Bayesian regression to sample the distribution of linear models that are compatible with this multivariate normal distribution, to obtain a statistically efficient estimate of $D^*$ and an accurate estimate of the associated statistical uncertainty.
\end{abstract}

@@

\section*{Significance Statement}
Quantifying diffusion in liquid and solid-state systems is fundamental to our understanding of their properties—without the ability to accurately measure diffusion coefficients, robust insight into these systems is impossible.
Here, we present a statistically efficient approach to estimating self-diffusion coefficients from single molecular dynamics simulations, using approximate Bayesian regression.
Our method provides estimates of self-diffusion coefficients with significantly less uncertainty than textbook ordinary-least-squares or weighted-least-squares methods, allowing for more efficient predictions of diffusion behaviour from simulations at greatly reduced overall computational cost, and also gives an accurate estimate of the variance in the predicted value, allowing for meaningful downstream comparison of data.

@@

Mass transport is a fundamental physical process that is central to our understanding of fluids~\cite{sendner_interfacial_2009,ShimizuEtAl_PhysChemChemPhys2015,ghoufi_ultrafast_2016} and plays a critical role in biochemical systems~\cite{maccmmon_dynamics_1977,RobertsonEtAl_ProcNatlAcadSci2006} and solid-state devices, such as batteries, fuel cells, and chemical sensors~\cite{eames_ionic_2015,morgan_understanding_2011,walsh_taking_2018}.
Molecular dynamics simulations are widely used to study microscopic transport processes, as they give direct insight into atomic-scale transport mechanisms and can be used to calculate macroscopic transport coefficients~\cite{morgan_relationships_2014,morgan_mechanistic_2021,poletayev_defect_2022,klepis_long_2009,wang_application_2011,ZelovichEtAl_ChemMater2019}.
These transport coefficients are formally defined in terms of ensemble averages.
Dynamical simulations, however, sample the full ensemble space stochastically, and parameters derived from simulation data, therefore, are estimates of the true parameter of interest.
The statistical uncertainty associated with such estimates depends on the details of the simulation---e.g., size and timescale---and on the choice of estimation method.
An optimal estimation method will minimise the uncertainty in the computed quantity, i.e., it will have high statistical efficiency, and will also allow this uncertainty to be accurately estimated.

One commonly used parameter for quantifying atomic-scale mass transport is the self-diffusion coefficient, $\D$, which describes diffusion in the absence of a chemical potential gradient.
$\D$ is related to the ensemble-average mean squared displacement (MSD), $\MSD{t}$, via the Einstein relation~\cite{einstein_uber_1905,helfand_transport_1960},
\begin{equation}
    \D = \lim_{t\to\infty}\frac{\MSD{t}}{6t},
    \label{equ:einstein}
\end{equation}
where $t$ is elapsed time.
Because numerical simulations are finite in time and space, MSDs obtained from simulation data always differ from the true ensemble average MSD.
One can, however, compute an estimate of the self-diffusion coefficient, $\Dest$, by fitting a linear model to the observed MSD and using the gradient of this fitted model in place of $\MSD{t} / t$ in \cref{equ:einstein}.

The simplest approach to fitting a linear model to MSD data from simulation is ordinary least squares regression (OLS).
OLS gives analytical expressions for the ``best fit'' regression coefficients (the slope and intercept) and their respective uncertainties, making it easy to implement and quick to perform.
OLS, however, is appropriate only for data that are both statistically independent and identically distributed.
Neither of these conditions holds for MSD data obtained from simulation, which instead are serially correlated and usually have unequal variances.
As a consequence, OLS is statistically inefficient, giving a relatively large statistical uncertainty in $\Dest$.
Furthermore, the textbook OLS expression for the uncertainty in $\Dest$ significantly underestimates the true uncertainty in this estimate~\cite{UslerEtAl_JComputChem2023}.
This underestimated uncertainty may give overconfidence in the accuracy of values of $\D$ estimated using OLS, and propagating these uncertainties in any downstream analyses may result in faulty inferences.
While the uncertainty associated with OLS estimates of $\D$ can, in principle, be accurately estimated by directly sampling over multiple repeated simulations, this approach can greatly increase the total computational cost and is often impractical.

Here, we describe an approximate Bayesian regression method for estimating $\D$ with near-maximal statistical efficiency while accurately estimating the corresponding statistical uncertainty using data from a single simulation.
We model the statistical population of simulation MSDs as a multivariate normal distribution, using an analytical covariance matrix derived for an equivalent system of freely diffusing particles, with this covariance matrix parameterised from the observed simulation data.
We then use Markov-chain Monte Carlo to sample the posterior distribution of linear models compatible with this multivariate normal model.
The resulting posterior distribution provides an efficient estimate for $\D$ and allows the associated statistical uncertainty in $\Dest$ to be accurately quantified.
This method is implemented in the open-source Python package \textsc{kinisi}~\cite{mccluskey_kinisi_2022}.

\section{Background}

For a simulation of equivalent particles, the observed mean squared displacement as a function of time, $\oMSDs{t}$, can be computed as an average over equivalent particles and time origins:
\begin{equation}
  \oMSDs{t} = \frac{1}{N(t)}\sum^{N(t)}_{j=1}{\left[\Delta\mathbf{r}_j(t)\right]}^2,
  \label{equ:observed_msd}
\end{equation}
where $N(t)$ is the total number of observed squared-displacements at time $t$.
The resulting observed MSD is a vector, $\oMSD$, with individual elements $\oMSDi$.
Each element of this vector differs from the true ensemble-average MSD for that time by some unknown amount.
Fitting a linear model to $\oMSD$ gives an estimated self-diffusion coefficient, $\Dest$, which again differs from the true self-diffusion coefficient, $\D$, by some unknown amount.

Performing repeated simulations starting from different random seeds or with different histories will produce a set of replica trajectories, where each trajectory gives a different, statistically equivalent, observed MSD.
The set of all possible replica trajectories defines a population of hypothetical observed MSDs, and the MSD obtained from any one trajectory can be considered a random sample, $\bm{X}$, drawn from the multivariate probability distribution that describes this population, i.e, $\bm{X} \sim \prob{\oMSD}$.
Each potential MSD sample could, in principle, be fitted to a linear model to obtain a corresponding estimate for the self-diffusion coefficient; $\bm{X} \mapsto \Dest$.
The population of all such estimates, therefore, defines a probability distribution $\prob{\Dest}$.
The estimated diffusion coefficient obtained from a single simulation corresponds to a random sample drawn from this distribution, while the uncertainty in $\Dest$ is described by the shape of the full distribution $\prob{\Dest}$.

@@

For data that are both statistically independent and identically normally distributed, ordinary least squares regression (OLS) is unbiased and statistically efficient, and gives accurate estimates of the uncertainties in the resulting regression coefficients.
MSD data obtained from simulation, however, are neither statistically independent nor identically distributed.
The variances, $\var{\oMSDi}$, are correlated, since the displacement of each particle at time $t+\Delta t$ is necessarily similar to its displacement at time $t$, and hence, $\oMSDs{t}$ is similar to $\oMSDs{t+\Delta t}$.
These variances are also typically unequal---the data are heteroscedastic \cite{smith_random_1996,he_statistical_2018,UslerEtAl_JComputChem2023}. 
Because the key assumptions of the OLS method are not valid for MSD data, OLS gives statistically inefficient estimates of $\D$, while the estimated regression uncertainties obtained from the standard OLS statistical formulae significantly underestimate the true uncertainty in $\prob{\Dest_\mathrm{OLS}}$ (\cref{fig:glswlsols}a).

\begin{figure}
    \centering
    \resizebox{\columnwidth}{!}{\includegraphics*{figure1.pdf}}
    \caption{
        Example distributions of estimated self-diffusion coefficients, $\Dest$, calculated using (a) ordinary least squares (OLS), (b) weighted least squares (WLS), and (c) generalised least squares (GLS),
        from MSD data from \num{4096} individual simulations of \num{128} particles undergoing a \SI{128}{step} 3D lattice random walk, with a step size chosen so that the true diffusion coefficient $\D = 1$.
        In each panel, the grey curve shows the best-fit normal distribution for the simulation data, the upper horizontal bar shows the standard deviation of this distribution, and the lower horizontal bar shows the average estimated standard deviation given by the analytical expression for $\sigma[\prob{\Dest}]$ for each regression method.}
    \label{fig:glswlsols}
    \script{glswlsols.py}
\end{figure}

Some improvement can be made by using weighted least squares (WLS) (\cref{fig:glswlsols}b), where the residual for each observed MSD value is weighted by the reciprocal of its variance, $1/(\var{\oMSDi})$.
Like OLS, WLS is an unbiased estimator, and for heteroscedastic data it has higher statistical efficiency than OLS.
WLS still neglects correlations in $\oMSD$, however, and is therefore statistically inefficient, and the WLS estimated uncertainties for the regression coefficients still underestimate the true uncertainty in $\prob{\Dest_\mathrm{WLS}}$.

To optimally estimate the true ensemble-average MSD, and hence $\D$, from simulation data, it is necessary to account for both the changing variance and correlation structure of $\oMSD$.
Within the framework of linear regression, this can be achieved using generalised least squares (GLS).
GLS gives estimated regression coefficients, $\widehat{\beta}$, via
\begin{equation}
    \widehat{\beta} = \left(\modelmatrix^{\top}\mathbf{\Sigma}^{-1}\modelmatrix\right)^{-1}\modelmatrix^{\top}\mathbf{\Sigma}^{-1}\oMSD,
    \label{equ:gls}
\end{equation} 
where $\modelmatrix$ is the model matrix $\begin{bmatrix}\mathbf{1} & \bm{t}\end{bmatrix}$, with $\bm{t}$ the vector of observed times, and $\mathbf{\Sigma}$ is the covariance matrix for the observed MSD values.
For correlated heteroscedastic data, such as MSD data, GLS offers the theoretical maximum statistical efficiency---it achieves the Cram\'er--Rao bound~\cite{cramer_mathematical_1946,rao_information_1945,rao_selected_1994,darmois_sur_1945,aitken_on_1942}---and provides accurate analytical estimates of the uncertainty in the predicted regression coefficients (\cref{fig:glswlsols}c).

An alternative method for estimating the ensemble-average MSD, and thus $\Dest$, from simulation data is Bayesian regression.
Like GLS, Bayesian regression can take into account both the changing variance and the correlation structure inherent in the data.
Rather than providing a singular ``best-fit'' estimate like GLS, Bayesian regression produces a posterior joint probability distribution for the regression coefficients.
The mean of this distribution serves as a point estimate of the coefficients and, in the absence of additional prior information, is equal to the estimate obtained from GLS, while the spread of the distribution quantifies the uncertainty in these estimates.
For data that are both heteroscedastic and correlated, such as MSD data from simulations, Bayesian regression, like GLS, is formally fully statistically efficient.

The estimation of $\D$ from some observed MSD data, $\oMSD$, using Bayesian regression, proceeds by computing the posterior probability distribution $\prob{\model|\oMSD}$ for a linear model $\model = 6\D \bm{t} + c$, where $\D$ and $c$ are parameters to be estimated.
This posterior distribution is described by Bayes' theorem,
\begin{equation}
    \prob{\model|\oMSD} = \frac{\prob{{\oMSD|\model}}\prob{\model}}{\prob{\oMSD}},
    \label{equ:bayes}
\end{equation}
where $\prob{\oMSD|\model}$ is the probability of observing data $\oMSD$ given model $\model$, often described as the ``likelihood'', and $\prob{\oMSD}$ is the marginal probability of the observed data $\oMSD$.
Integrating over $\prob{\model|\oMSD}$ with respect to $c$ yields the marginal posterior distribution $\prob{\D|\oMSD}$, from which the best point-estimate $\Dest$ and distribution variance $\varest{\Dest}$ can be computed.

Given a sufficiently large number of observed squared displacements at each time $t$, the central limit theorem applies, and $\oMSD$ can be considered a sample from a multivariate normal distribution with log-likelihood
\begin{equation}
    \begin{aligned}{}
      \ln \prob{\oMSD|\model} = -\frac{1}{2}\big[ & \ln(\left|\mathbf{\Sigma}\right|) + {(\oMSD - \model)}^{\!\top}\mathbf{\Sigma}^{-1}(\oMSD - \model) \\ 
      & + k \ln(2\pi)\big],
    \end{aligned}
    \label{equ:loglike}
\end{equation}
where $\mathbf{\Sigma}$ is the observed MSD covariance matrix and $k$ is the length of the vector $\oMSD$, i.e., the number of time intervals for which we have observed MSD data.
Providing that this likelihood function can be calculated, one can compute the posterior distribution $\prob{\model | \oMSD}$ via \cref{equ:bayes}, to obtain an optimally efficient point-estimate for $\D$ and a complete description of the associated uncertainty in $\Dest$.

\section{Approximating $\bm{\Sigma}$ from simulation data}

The practical application of Bayesian regression or GLS requires the covariance matrix for the observed MSD, $\mathbf{\Sigma}$, which is generally unknown.
To proceed, we approximate $\mathbf{\Sigma}$ with a model covariance matrix, $\mathbf{\Sigma^\prime}$, with a known analytical form, that we parameterise from the available simulation data.
Providing the correlation structure of $\mathbf{\Sigma^\prime}$ is similar to that of $\mathbf{\Sigma}$, this model correlation matrix can be used in approximate Bayesian or GLS schemes to estimate the ensemble-average MSD, and hence $\D$, with high efficiency and accurate estimated uncertainties.

We model the covariance matrix for the observed MSD from a given simulation using the covariance matrix for the MSD of an equivalent system of freely diffusing particles, $\mathbf{\Sigma^\prime}$.
We note that estimating \Dest\ by fitting a linear model to observed MSD data implicity assumes that these data sample the long-time limit where the Einstein relation (\cref{equ:einstein}) is valid.
In this long-time diffusive regime, all systems of mobile particles are statistically equivalent under rescaling by \D, and hence have the same MSD covariance structure $\mathbf{\Sigma}$.

For observed MSDs computed by averaging over non-overlapping time windows, the covariance matrix $\mathbf{\Sigma^\prime}$, in the long-time limit, has elements (see SI)
\begin{equation}
  \Sigma^\prime\left[\oMSDi, \oMSDj\right]= \Sigma^\prime\left[\oMSDj, \oMSDi\right] =
  \var{\oMSDi} \frac{\nind{i}}{\nind{j}},\hspace{1em} \forall\,i \leq j,
  \label{equ:cvv}
\end{equation} 
where $\var{\oMSDi}$ are the time-dependent variances of the observed MSD, and $\nind{i}$ is the total number of non-overlapping observed squared-displacements for time-interval $i$.
We estimate the variances $\var{\oMSDi}$ using the standard result that the variance of the mean of a sample scales inversely with the number of independent constituent observations.
Specifically, we compute an estimated variance $\varest{\oMSDi}$ by rescaling the observed variance of the squared displacement for time interval $i$ by the number of numerically-independent contributing sub-trajectories, $\nind{i}$, which is given by the number of mutually non-overlapping time windows of length $i$ multiplied by the number of mobile particles, summed over all simulations used to compute the MSD;
\begin{equation}
  \varest{\oMSDi} = \frac{1}{\nind{i}}\var{\Delta r_i^2}.
  \label{equ:varestMSD}
\end{equation}
By rescaling by the number of numerically-independent contributing sub-trajectories, rather than by the total number of observed squared displacements for time window $i$, we account for correlations between the squared displacements of each particle computed from overlapping time windows (further details are provided in the SI).
